% $Id: phd-thesis-proposal.tex,v 1.5 2008/10/22 12:32:30 rlunsford Exp $
%
\documentclass [11pt] {report}
\makeatletter
\let\orig@xfloat\@xfloat
\makeatother

\usepackage{lscape}		% Allows tables and figures in landscape mode
\usepackage{ohsu-thesis-proposal_3-2012}		% This package is the ohsu thesis style stuff
% \brokenpenalty=100		% uncomment this to see bad word breaks at EOP
% \usepackage{colortab}
\usepackage{epsfig}			% This package allows encapsulated postscript
\usepackage{subfigure}		% This package allows subfigures
\usepackage{graphicx}
\usepackage{footmisc}
\usepackage{amssymb}
%\usepackage[style=acm]{biblatex}
%\renewbibmacro{in:}{%
% \ifentrytype{article}{}{%
% \printtext{\bibstring{in}\intitlepunct}}}
%\addbibresource{nm.bib}

\usepackage{etoolbox}
\makeatletter
\let\@xfloat\orig@xfloat
\patchcmd{\@xfloat}{\columnwidth}{\columnwidth\def\baselinestretch{\@ne}}{}{}
\makeatother

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{url}
\usepackage{multirow}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{xr}

\usepackage[T1]{fontenc}

% \usepackage{algorithm}
% \usepackage[noend]{algpseudocode}
% \usepackage{url}
% \usepackage{amsfonts}
% \usepackage{amsmath}
% \usepackage{authblk}
% \usepackage{rotating}
% \usepackage{xr}



\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1} 
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

\newcommand{\ignore}[1]{}
\newcommand{\rl}[1]{}
\newcommand{\peter}[1]{}
\newcommand{\hints}[1]{}
%\newcommand{\todo}[1]{ **NOW: \emph{#1}**}
%\newcommand{\todolater}[1]{ **LATER: \emph{#1}**}
\newcommand{\todo}[1]{}
\newcommand{\todolater}[1]{}

\makeatletter
\let\latex@xfloat=\@xfloat
\def\@xfloat #1[#2]{%
 \latex@xfloat #1[#2]%
 \def\baselinestretch{1}
 \@normalsize\normalsize
 \normalsize
}
\makeatother

\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\degree}{\ensuremath{^\circ}}

%\includeonly{breakpoint-analysis,future-work}

\begin{document}

% Specify your title here. The "titleline" field should have no line breaks
% in it.
%
\title{Detecting and Analyzing Genomic Structural Variation using Distributed Computing}
\titleline{Detecting and Analyzing Genomic Structural Variation using Distributed Computing}

% Your name (full name, no initials)
%
\author{Christopher Whelan}

% Your previous degree. If you have more than one, use line breaks e.g.,
% \predegrees{degree1 \\ degree2}
%
\prevdegrees{A.B. Computer Science, Harvard University, 1997}
	
	% Heirarchy info
	\discipline{Computer Science \& Engineering}
	\division{Center for Spoken Language Understanding}
	\dept{Institute on Development and Disability}
	\school{School of Medicine}
	\university{Oregon Health \& Science University}

% The following information is for the signature page.
% Note that the definition for principal adviser uses two fields.
% This was needed so that the adviser's name could be placed on the
% abstract page without his title.
%
 %\foursigstrue | \fivesigstrue but don't define BOTH to be true!!
\fivesigstrue
\principaladviser{Dr. Kemal S\"onmez}{Associate Professor}
\secondreader{Dr. Brian Roark \\Research Scientist, Google, Inc.}
\firstreader{Dr. Izhak Shafran \\Associate Professor}
\thirdreader{Dr. Lucia Carbone \\Assistant Professor, Dept. of Behavioral Neuroscience}
\fourthreader{Dr. Steven Bedrick \\Assistant Professor}

% This is the expected date that the board of trustee's sign your thesis. They
% meet four times a year and the dates can be obtained from either the Office
% of Admissions and Records (Margaret Day) or from the department secretary.
% In 1991 they were meeting the fourth Monday of each quarter, e.g.,
% January, April, July, October.
\grantdate{February}{2014}

% If you want to copyright your thesis/dissertation remove the line below.
\copyrightfalse		% True by default

% The year of the copyright; usually same as the date the Board of Trustees
% sign the thesis. This won't be printed if \copyrightfalse
\copyrightyear{2014}

% The prefatory section of this style is normally used for the Dedication
% and/or Acknowledgements. You might not want to put this into your thesis
% until the end, to avoid generating it when you are still in the writing
% stage.
%
% The below creates the title page, etc.
\beforepreface

% This is a sample dedication. It should be vertically centered in the page.
% Since the style format doesn't do it for you automatically, you can use
% the following technique.
\ignore{\prefacesection{Dedication}
\vfill
\begin{center}
\end{center}
\vfill
}
% This is a sample acknowledgements page. As in the dedication, it should
% be centered vertically in the page.
%
\prefacesection{Acknowledgments}
\vfill
\indent This dissertation would not be possible without my co-advisors, Kemal S\"onmez and Lucia Carbone. Kemal has been an excellent mentor and a great friend. He helped me to develop many ideas that went into this dissertation. Of all of his valuable contributions, I am most thankful for his ability to point me in new directions and re-energize me when I was discouraged or had painted myself into a corner. 

Lucia has also been a terrific role model and an inspiring scientific leader. The day I was given the chance to collaborate with her and her lab was my luckiest moment at OHSU. Lucia has taught me most of the biology I know, given me the opportunity to work on tremendously exciting projects, and gone out of her way many times to introduce me to members of the genomics community; in particular, I owe a great debt of gratitude to her for bringing me annually to the Biology of Genomes meeting at Cold Spring Harbor. Computational work cannot have real value without actual biological data and real scientific questions to work on and learn from, both of which Lucia provided. Of course, Lucia also introduced me to the problem of structural variation detection; this dissertation could not exist without her. I would also like to thank the other members of the Carbone Lab, who have been a pleasure to work with and learn from: Larry Wilhelm, Josh Meyer, Nathan Lazar, Liz Terhune, and Kim Nevonen.

I would also like to thank Brian Roark and Zak Shafran for all of the great help they've given me. Brian is a fabulous teacher and always provided excellent feedback and made himself available for questions or just to talk. Cloudbreak started as a project in a course taught by Zak and Richard Sproat on problem solving with large clusters, which provided me with the core idea that eventually turned into this dissertation. Zak has done a great deal of other things for me as well: providing me with ideas to use in my research; inviting me to explore new topics in machine learning with him and his group; and last but not least acquiring and helping to configure and administer the Hadoop compute clusters at CSLU, without which this work would not be possible. On that last note, I would also like to thank cluster administrators who have put up with my sometimes taxing usage of our compute resources, including Rob Stites and Jason Brooks.

I would also like to acknowledge members of the broader genomics community who have provided feedback and encouragement for my projects, including Bob Handsaker, Steve McCarroll, and Ben Raphael.

Finally, my greatest thanks go to my wife Sarah and my son Colin, who displayed tremendous patience in putting up with their husband and father's graduate-school lifestyle. Sarah's support, encouragement, and love have been an integral component of this work, without which it could never have happened.
\vfill

% Uncomment the line below if you don't want a list of tables to be printed.
%\tablespagefalse

% Uncomment the line below if you don't want a list of figures to be printed.
%\figurespagefalse

% \afterpreface generates the table of contents, list of tables (optional),
% and list of figures (optional).
\afterpreface

% Abstract
\include{abstract}

% This is where the main body of the thesis starts
%
\body
%\setcounter{chapter}{6}

\include{introduction}

\include{background}

\include{breakpoint-analysis}

\include{related_work}

\include{framework}

\include{cloudbreak-implementation}

\include{cloudbreak-evaluation}

\include{crf}

\include{future-work}


% The library likes numbers rather than the `alpha' style.
\bibliographystyle{acm}
% Single space the bibliography to save space.
\begin{singlespace}
\bibliography{thesis}
\end{singlespace}
%\printbibliography

% The appendices are optional, but they must follow the references.
%\appendix

% This section (the final one) contains a biographical note.
%
%\vita
%This section contains a biographical note.
%The following information (in essay form) should be contained in this section:
%place of birth; date of birth; schools attended; degrees awarded; areas of
%special interest; relevant profession experience; awards and honors; list
%of publications.

\end{document}
